Supplement: Efficient inference, potential, and limitations of site-specific substitution models

### 2020-01-18 manuscript/supplement.tex
typos, changed position label from a to b

\title{Supplement: Potential and limits of site-specific substitution model inference.}
\author{Vadim Puller$^{1,2}$ Pavel Sagulenko$^{3}$, Richard A.~Neher$^{1,2}$}
\affiliation{
$^{1}$Biozentrum, University of Basel, Klingelbergstrasse 50/70, 4056 Basel, Switzerland\\
$^{2}$SIB Swiss Institute of Bioinformatics, Basel, Switzerland\\
$^{3}$Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}

@@

At the maximum, the derivatives of $\lh$ with respect to each of the model parameters should vanish when averaged over all internal site assignments.
The derivative with respect to $\mu^b$ was is presented in Eq.~12 of the main text.
Here, we present the corresponding derivations for $\eqp^b_n$ and $W_{nm}$.

To facilitate the algebra below, we summarize the short branch length assumption and the associated linear approximations here:
\begin{equation}
	\left( e^{Q t}\right)_{ij} = \delta_{i,j} + \sum_{m=1}^\infty \frac{t^m}{m!}\left( Q^m\right)_{ij} \approx
	\begin{cases}
	1 - t\mu^b \sum_{k\neq i} \eqp_k^b W_{ki} & i=j \\
	t\mu^b \eqp^b_i W_{ik} & i\neq j
	\end{cases}
\end{equation}
The derivatives of $\left( e^{Q t}\right)_{ij}$ with respect to the different model parameters are therefore
\begin{eqnarray}
	\frac{d}{d\mu^b} \left( e^{Q^a t}\right)_{ij} &\approx & \delta_{ab}
	\begin{cases}
	-t\sum_{k\neq i} \eqp_k^b W_{ki} & i=j \\
	t \eqp^b_i W_{ik} & i\neq j
	\end{cases} \\
	\frac{d}{d\eqp_n^b} \left( e^{Q^a t}\right)_{ij} &\approx & \delta_{ab}
	\begin{cases}
	-t\mu^b (1-\delta_{i,n})W_{ni} & i=j \\
	 t\mu^b \delta_{i,n}W_{nk} & i\neq j
	\end{cases}\\
	\frac{d}{dW_{mn}} \left( e^{Q^a t}\right)_{ij} &\approx &
	\begin{cases}
	-\delta_{n,j}(1-\delta_{m,i})t \mu^a \eqp^a_m  & i=j \\
	 \delta_{m,i}\delta_{n,j} t\mu^a \eqp^a_m & i\neq j
	\end{cases}
\end{eqnarray}
Since the matrix $W_{ij}$ is symmetric, the derivative with respect to $W_{ij}$ and $W_{ji}$ are the same.
We will account for this symmetry below explicitly and refrain from carrying the symmetric term through lengthy derivations.
Note that the derivatives with respect to $W_{ii}$ vanish as required.

\section*{Update rule for rates $\mu^b$}
Only site $b$ will contribute to the derivative of $\lh$ with respect to $\mu^b$.
We will therefore drop all indices referring to the position and re-instantiate the index later.
As discussed in the main text, we will separate the sum over branches $\beta$ into those where sequence is the same of parent and child ($s_c=s_p$) and those where the sequence differs ($s_c\neq s_p$).
\begin{equation}
\begin{split}
	\frac{d}{d \mu} \lh (\{\vec{s}\} |T, \mat{Q}) &
	 \approx  -\sum_{\beta\in T,s_c=s_p} \frac{t_\beta \sum_{k\neq s_c} \eqp_{k} W_{k s_c}}{1 - t_\beta \mu \sum_{k\neq s_c} \eqp_k W_{ks_c}}
	 + \sum_{\beta\in T,s_c\neq s_p} \frac{t_\beta \eqp_{s_c} W_{s_cs_p}}{t_\beta \mu \eqp_{s_c} W_{s_c s_p}}\\
	 & \approx  -\sum_{\beta\in T,s_c=s_p} t_\beta \sum_{k\neq s_c} \eqp_{k} W_{k s_c} + \sum_{\beta\in T,s_c\neq s_p} \frac{1}{\mu} = - \sum_{j, k\neq j} \eqp_k W_{kj} \tau_j + \sum_{i\neq j} n_{ij}/\mu
\end{split}
\end{equation}
Here, $\tau_j$ is the sum of all branch lengths on which $s_c=j$ and $n_{ij}$ is the number of times the sequence changes from $j$ to $i$ somewhere along the tree.
Setting this expression to 0 and re-instantiating the site index $b$, we get
\begin{equation}
	\mu^b = \frac{\sum_{i\neq j} n^b_{ij}}{\sum_{j, k\neq j} \eqp^b_k W_{kj} \tau^b_j}
\end{equation}

\section*{Update rule for frequencies $\eqp_i^b$}
As above, only site $b$ will contribute to the derivative of $\lh$ with respect to $\eqp^b_n$.
We will therefore drop all indices referring to the position and re-instantiate the index later.
\begin{equation}
\begin{split}
	\frac{d}{d \eqp_n} \lh (\{\vec{s}\} |T, \mat{Q}) & \approx \frac{\delta_{s_0,n}}{\eqp_n}
	-\sum_{\beta\in T,s_c=s_p} \frac{t_\beta\mu (1-\delta_{n,s_c})W_{n s_c}}{1 - t_\beta\mu \sum_{k\neq s_c} \eqp_k W_{ks_c}}
	 + \sum_{\beta\in T,s_c\neq s_p} \delta_{n,s_c}\frac{t_\beta\mu W_{ns_p}}{t_\beta \mu \eqp_{s_c} W_{s_c s_p}}\\
	 & \approx 	\frac{\delta_{s_0,n}}{\eqp_n}-\sum_{\beta\in T,s_c=s_p} t_\beta\mu (1-\delta_{s_c, n})W_{s_c n}
	 + \sum_{\beta\in T,s_c=n, s_c\neq s_p} \frac{1}{\eqp_{n}}\\
	 & = 	\frac{\delta_{s_0,n}}{\eqp_n}-\sum_{j\neq n} \tau_j \mu W_{j n}
	 + \sum_{j\neq n}\frac{n_{nj}}{\eqp_{n}}
\end{split}
\end{equation}
Reinstantiating the position index $b$, we find
\begin{equation}
	\eqp^b_n = \frac{\delta_{n,s^b_0}+\sum_j n^b_{nj}}{\sum_{j\neq n} \mu W_{nj}\tau^b_j}
\end{equation}

@@

\section*{Implementation of the constraint $\sum_i \eqp_i^b=1$}
Equilibrium frequencies need to be normalized to one and constraints of this nature are typically accounted for by Lagrange parameters
\begin{equation}
	\frac{d}{d \eqp_n} \left[\lh (\{\vec{s}\} |T, \mat{Q}) + \lambda^b \sum_i \eqp^b_i \right] = 0
\end{equation}
which results in a modified extremal condition
\begin{equation}
	0 = \delta_{s^b_0,n}-\eqp_{n}\sum_{j\neq n} \tau^b_j \mu^b W_{j n} + \sum_{j\neq n} n^b_{nj} + \eqp^b_{n}\lambda^b
\end{equation}
Summing this condition over $n$, we find
\begin{equation}
	0 = 1 - \sum_{i,j; i \neq n}  \mu^b \tau^b_j \eqp^b_{i}W_{j i} + \sum_{i,j; i\neq j} n^b_{ij} + \lambda^b
\end{equation}
At the extremum of $\mu^v$, we have $\sum_{i,j; i \neq n} \mu^b \tau^b_j \eqp^b_{i}W_{j i} = \sum_{i,j; i\neq j} n_{ij}$ and therefore $\lambda^b=-1$.
With the constraint $\sum_i \eqp_i^b=1$, the update rule therefore becomes
\begin{equation}
	\eqp^b_n = \frac{\delta_{n,s^b_0}+\sum_j n^b_{nj}}{1 + \sum_{j\neq n} \mu W_{nj}\tau^b_j}
\end{equation}
In practice it is often helpful to explicitly enforce normalization after each iteration.

\section*{Regularization}
In absence of data, i.e., when there are very few mutations at a particular site we need to regularize the problem.
To this end, we assume a Dirichlet prior of the form
\begin{equation}
\prod_i \left(\eqp^b_i\right)^c
\end{equation}
The derivative of the logarithm of this is $c/\eqp^b_i$ and adding this term to the update rule results in
\begin{equation}
	\eqp^b_n = \frac{\delta_{n,s^b_0} + c +\sum_j n^b_{nj}}{1 + qc + \sum_{j\neq n} \mu W_{nj}\tau^b_j}
\end{equation}
where $q$ is the size of the alphabet.
In an analogous fashion, we add a weak regularization to the update rules of $\mu^b$ and $W_{ij}$.

\section*{Summation over unobserved states}
So far, we have derived conditions that maximize the likelihood as a function of parameters of the substitution model for a fixed assignment of sequences at internal nodes.
In practice, the internal nodes are unknown and can either be fixed to the most-likely state or summed over.
Since the approximate maximum likelihood conditions are linear, this average over unknown ancestral states simply amounts to replacing $\tau^b_i$ and $n^b_{ij}$ by their averages which can be efficiently computed by standard recursive algorithms.

>>>>

\title{Supplement: Efficient inference, potential, and limitations of site-specific substitution models}
\author{Vadim Puller$^{1,2}$ Pavel Sagulenko$^{3}$, Richard A.~Neher$^{1,2}$}
\affiliation{
$^{1}$Biozentrum, University of Basel, Klingelbergstrasse 50/70, 4056 Basel, Switzerland\\
$^{2}$SIB Swiss Institute of Bioinformatics, Basel, Switzerland\\
$^{3}$Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}

@@

At the maximum, the derivatives of $\lh$ with respect to each of the model parameters should vanish when averaged over all internal site assignments.
To facilitate the algebra below, we summarize the short branch length assumption and the associated linear approximations here:
\begin{equation}
	\left( e^{Q t}\right)_{ij} = \delta_{i,j} + \sum_{m=1}^\infty \frac{t^m}{m!}\left( Q^m\right)_{ij} \approx
	\begin{cases}
	1 - t\mu^a \sum_{k\neq i} \eqp_k^a W_{ki} & i=j \\
	t\mu^a \eqp^a_i W_{ik} & i\neq j
	\end{cases}
\end{equation}
The derivatives of $\left( e^{Q t}\right)_{ij}$ with respect to the different model parameters are therefore
\begin{eqnarray}
	\frac{d}{d\mu^a} \left( e^{Q^a t}\right)_{ij} &\approx & \delta_{ab}
	\begin{cases}
	-t\sum_{k\neq i} \eqp_k^a W_{ki} & i=j \\
	t \eqp^a_i W_{ik} & i\neq j
	\end{cases} \\
	\frac{d}{d\eqp_n^a} \left( e^{Q^a t}\right)_{ij} &\approx & \delta_{ab}
	\begin{cases}
	-t\mu^a (1-\delta_{i,n})W_{ni} & i=j \\
	 t\mu^a \delta_{i,n}W_{nk} & i\neq j
	\end{cases}\\
	\frac{d}{dW_{mn}} \left( e^{Q^a t}\right)_{ij} &\approx &
	\begin{cases}
	-\delta_{n,j}(1-\delta_{m,i})t \mu^a \eqp^a_m  & i=j \\
	 \delta_{m,i}\delta_{n,j} t\mu^a \eqp^a_m & i\neq j
	\end{cases}
\end{eqnarray}
Since the matrix $W_{ij}$ is symmetric, the derivative with respect to $W_{ij}$ and $W_{ji}$ are the same.
We will account for this symmetry below explicitly and refrain from carrying the symmetric term through lengthy derivations.
Note that the derivatives with respect to $W_{ii}$ vanish as required.

\section*{Update rule for rates $\mu^a$}
Only site $a$ will contribute to the derivative of $\lh$ with respect to $\mu^a$.
We will therefore drop all indices referring to the position and re-instantiate the index later.
As discussed in the main text, we will separate the sum over branches $\beta$ into those where sequence is the same of parent and child ($s_c=s_p$) and those where the sequence differs ($s_c\neq s_p$).
\begin{equation}
\begin{split}
	\frac{d}{d \mu} \lh (\{\vec{s}\} |T, \mat{Q}) &
	 \approx  -\sum_{\beta\in T,s_c=s_p} \frac{t_\beta \sum_{k\neq s_c} \eqp_{k} W_{k s_c}}{1 - t_\beta \mu \sum_{k\neq s_c} \eqp_k W_{ks_c}}
	 + \sum_{\beta\in T,s_c\neq s_p} \frac{t_\beta \eqp_{s_c} W_{s_cs_p}}{t_\beta \mu \eqp_{s_c} W_{s_c s_p}}\\
	 & \approx  -\sum_{\beta\in T,s_c=s_p} t_\beta \sum_{k\neq s_c} \eqp_{k} W_{k s_c} + \sum_{\beta\in T,s_c\neq s_p} \frac{1}{\mu} = - \sum_{j, k\neq j} \eqp_k W_{kj} \tau_j + \sum_{i\neq j} n_{ij}/\mu
\end{split}
\end{equation}
Here, $\tau_j$ is the sum of all branch lengths on which $s_c=j$ and $n_{ij}$ is the number of times the sequence changes from $j$ to $i$ somewhere along the tree.
Setting this expression to 0 and re-instantiating the site index $a$, we get
\begin{equation}
	\mu^a = \frac{\sum_{i\neq j} n^a_{ij}}{\sum_{j, k\neq j} \eqp^a_k W_{kj} \tau^a_j}
\end{equation}

\section*{Update rule for frequencies $\eqp_i^a$}
As above, only site $a$ will contribute to the derivative of $\lh$ with respect to $\eqp^a_n$.
We will therefore drop all indices referring to the position and re-instantiate the index later.
\begin{equation}
\begin{split}
	\frac{d}{d \eqp_n} \lh (\{\vec{s}\} |T, \mat{Q}) & \approx \frac{\delta_{s_0,n}}{\eqp_n}
	-\sum_{\beta\in T,s_c=s_p} \frac{t_\beta\mu (1-\delta_{n,s_c})W_{n s_c}}{1 - t_\beta\mu \sum_{k\neq s_c} \eqp_k W_{ks_c}}
	 + \sum_{\beta\in T,s_c\neq s_p} \delta_{n,s_c}\frac{t_\beta\mu W_{ns_p}}{t_\beta \mu \eqp_{s_c} W_{s_c s_p}}\\
	 & \approx 	\frac{\delta_{s_0,n}}{\eqp_n}-\sum_{\beta\in T,s_c=s_p} t_\beta\mu (1-\delta_{s_c, n})W_{s_c n}
	 + \sum_{\beta\in T,s_c=n, s_c\neq s_p} \frac{1}{\eqp_{n}}\\
	 & = 	\frac{\delta_{s_0,n}}{\eqp_n}-\sum_{j\neq n} \tau_j \mu W_{j n}
	 + \sum_{j\neq n}\frac{n_{nj}}{\eqp_{n}}
\end{split}
\end{equation}
Reinstantiating the position index $a$, we find
\begin{equation}
	\eqp^a_n = \frac{\delta_{n,s^a_0}+\sum_j n^a_{nj}}{\sum_{j\neq n} \mu W_{nj}\tau^a_j}
\end{equation}

@@

\section*{Implementation of the constraint $\sum_i \eqp_i^a=1$}
Equilibrium frequencies need to be normalized to one and constraints of this nature can be accounted for by Lagrange parameters $\lambda^a$ coupling to the constraint:
\begin{equation}
	\frac{d}{d \eqp_n} \left[\lh (\{\vec{s}\} |T, \mat{Q}) + \lambda^a \sum_i \eqp^a_i \right] = 0
\end{equation}
The constraint results in a modified extremal condition
\begin{equation}
	0 = \delta_{s^a_0,n}-\eqp_{n}\sum_{j\neq n} \tau^a_j \mu^a W_{j n} + \sum_{j\neq n} n^a_{nj} + \eqp^a_{n}\lambda^a
\end{equation}
Summing this condition over $n$, we find
\begin{equation}
	0 = 1 - \sum_{i,j; i \neq n}  \mu^a \tau^a_j \eqp^a_{i}W_{j i} + \sum_{i,j; i\neq j} n^a_{ij} + \lambda^a
\end{equation}
At the extremum of $\mu^a$, we have $\sum_{i,j; i \neq n} \mu^a \tau^a_j \eqp^a_{i}W_{j i} = \sum_{i,j; i\neq j} n_{ij}$ and therefore $\lambda^a=-1$.
With the constraint $\sum_i \eqp_i^a=1$, the update rule therefore becomes
\begin{equation}
	\eqp^a_n = \frac{\delta_{n,s^a_0}+\sum_j n^a_{nj}}{1 + \sum_{j\neq n} \mu W_{nj}\tau^a_j}
\end{equation}
In practice it is often helpful to explicitly enforce normalization after each iteration.

\section*{Regularization}
When there are very few mutations at a particular site we need to regularize the problem to avoid overfitting and numerical instabilities.
To this end, we assume a Dirichlet prior of the form
\begin{equation}
\prod_i \left(\eqp^a_i\right)^c
\end{equation}
which drives frequency estimates away from the boundary for $c>0$.
The derivative of the logarithm of this prior is $c/\eqp^a_i$ and adding this term to the update rule results in
\begin{equation}
	\eqp^a_n = \frac{\delta_{n,s^a_0} + c +\sum_j n^a_{nj}}{1 + qc + \sum_{j\neq n} \mu W_{nj}\tau^a_j}
\end{equation}
where $q$ is the size of the alphabet.
In an analogous fashion, we add a weak regularization to the update rules of $\mu^a$ and $W_{ij}$.

\section*{Summation over unobserved states}
So far, we have derived conditions that maximize the likelihood as a function of parameters of the substitution model for a fixed assignment of sequences at internal nodes.
In practice, the internal nodes are unknown and can either be fixed to the most-likely state or summed over.
Since the approximate maximum likelihood conditions are linear, this average over unknown ancestral states simply amounts to replacing $\tau^a_i$ and $n^a_{ij}$ by their averages which can be efficiently computed by standard recursive algorithms.
